---
title: "DESeq2_CMC_Tutorial_Part3"
author: "Nickie Safarian"
date: "11/23/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In script 3, steps towards visualizing the selected Neuropeptides 
genes' counts and demographic features are described. The main questions is:

  - How does the expression of the select neuropeptides change with age?
   

## Load the packages

```{r}

library(tidyverse)

```


## Import the Neuropeptides list/ files

```{r}

# Vanessa's Reported NPs list
Age_NPs <- read.csv(file=file.path("/external/rprshnas01/kcni/nsafarian/CMC_Data_Analysis/Neuropeptides/Age_NPs.csv")) # 81 NPs

GO_NPs <- read.csv(file=file.path("/external/rprshnas01/kcni/nsafarian/CMC_Data_Analysis/Neuropeptides/GOSelected_NPs.csv")) # 69 NP genes

```
Note: The neuropeptide files have only gene's name, and my attempt here is to add 
ENSEMBL_gene_ID column to these files. This may help with the  process of merging Age_related_DEG file to the NPs list. 


## Add Ensembl.IDs column to the NPs' gene names 

```{r}

library( "biomaRt" )
ensembl <- useMart( "ensembl", dataset = "hsapiens_gene_ensembl" )
gene.map <- getBM(attributes = 
                    c("ensembl_gene_id", 'external_gene_name', "hgnc_symbol"),
                  filters = "hgnc_symbol", 
                  values = Age_NPs$Gene.Symbol, 
                  mart = ensembl) # 86 genes is listed, but the
                                  # Age_NPs has 81 Gene.Symbol!!! duplication issue:

# How many duplicate IDs are there?
sum(duplicated(gene.map$hgnc_symbol))  # 8 duplicated

# You may remove duplicates (I don't recommend):
# gene.map <- gene.map[!duplicated(gene.map$hgnc_symbol), ] #79 ids remain

# merge
Age_NPs.Full <- left_join(gene.map, Age_NPs, by=c('hgnc_symbol' = 'Gene.Symbol'))

# now the NPs' gene names has the Ensembl gene IDs column attached to it.

```


## Import the DESeq2 driven Age-related differentially expressed genes

```{r}

CMC_Age_DEGs <- read.csv(file=file.path("/external/rprshnas01/kcni/nsafarian/CMC_Data_Analysis/CommonMind.Data/DESeq2Object/DESeq2.results_for_Age.csv"))  

# set the rownames
CMC_Age_DEGs <- CMC_Age_DEGs %>% column_to_rownames(var="X") #27941 genes

```


## Subset DESeq2 driven Age-related differentially expressed genes for NPS

```{r}

select.genes <- CMC_Age_DEGs %>% subset(ensemblID %in% Age_NPs.Full$ensembl_gene_id) 
# 53 genes remain (Out of 81 NPs, only 53 matched with DESeq2 DEG results)
  
```


## construct the desired data matrix for visualization
In this matrix[x, y], rows will be samples and columns are grouping variables
as well as DEGs. 

### *Step 1. Import DEseq2 objects*

```{r}

dds <- readRDS(file=file.path("/external/rprshnas01/kcni/nsafarian/CMC_Data_Analysis/CommonMind.Data/DESeq2Object/dds_CMCdata_onlyControls.counts.more.than.10.Rds"))


vsd <- readRDS(file=file.path("/external/rprshnas01/kcni/nsafarian/CMC_Data_Analysis/CommonMind.Data/DESeq2Object/vst.Transformed_CMCdata_onlyControls_counts.more.than.10.Rds"))


```

### *2. Transform the count matrix and subset it for the select.genes*

```{r}

# Transform the data so samples are in rows and genes are in columns
geneCounts <- as.data.frame(t(assay(vsd)))

# Subset the transformed count matrix for only select.genes:
dt <- geneCounts[, colnames(geneCounts) %in% rownames(select.genes)] 
              # 451 Control Samples * 27941 genes 

```

### *3. Add grouping variables to the matrix*

```{r}

dt$Cohort <- dds@colData$Cohort
dt$Gender <- dds@colData$Reported_Gender
dt$PMI <- dds@colData$PMI
dt$RIN <- dds@colData$RIN
dt$Age <- dds@colData$ageOfDeath


# Rearrange, so the grouping variables are at the begining of the data
dt <- dt[, c(54: 58, 1:53)]

# Check the data
library(knitr)
kable(dt [1:6, 1:8])

# save
write.csv(dt, "CMC_vsd.data_with.variables_only.Controls_27941.DEG.csv")

```

#------------------------------------------------------------------------------

##  Q1- How does the expression of the select neuropeptides change with age?

### *Visualize the correlation between age and select.genes' expression* 

```{r}

# scatter plot for individual gene.count (across age)

ggplot(dt, aes(x=Age, y=ENSG00000175206.10)) + 
  geom_point() + 
  ggtitle("ENSG00000175206.10\n (expression across age)")+
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.text.y = element_text(size = 10),
        axis.title =element_text(size=10,face="bold"))+
  theme(plot.title = element_text(size = 8, face = "bold"))+
  scale_y_log10()

```


### *Box.Plot the expression of select.genes vs. other grouping/categorical variables (like gender, cohort, etc.)*

```{r}

# plot Function, how to make all plots in one run:

plots_fun <- function (df, x_var, y_var){
  ggplot(df, aes(x=.data[[x_var]], y=.data[[y_var]]))+
    geom_boxplot(outlier.shape=NA)+
    geom_jitter(shape=1, color="coral2", position=position_jitter(0.2))+
    ggtitle(paste0("...."), y_var)+
    labs(x= x_var, y=y_var)+
    theme(axis.text = element_text(size=10, face="bold"),
          axis.title = element_text(size=10, face="bold"))+
    theme_bw()
}


# call the function to make a list of plot 
NPs_plot_list <- colnames(dt [, -c(1:5)]) %>%  
  map (~ plots_fun (dt, colnames(dt)[1], .x))

# Note two things: 
     # 1) you need to remove the categorical variable columns before calling 
     # for the function (that's why i used "colnames(dt [, -c(1:5)])" , where 
     # column 1 to 5 are all categorical). 
     # 2) in the map() function the colnames(dt)[1] part asks to consider 
     # column 1 as the grouping variable for plotting counts. In this case it's 
     # cohorts. But for sure we can use Gender column too. 
     

# save
saveRDS(NPs_plot_list , "select_neuropeptides_plot_list.Rds")

```

